Geographically aware estimates of remotely sensed water properties for Chesapeake Bay

### 2022-12-09 manuscript/manuscript.tex
add sofia bio
use sofia's bio
rm highlighting

Unfortunately, direct observations alone are not adequate for many applications that require spatially extensive data such as \hl{Sea Level Rise (SLR)} forecasting and habitat suitability assessments \cite{hoodChesapeakeBayProgram2021}. For example, habitat suitability assessments are typically done on a gridded basis for the entirety of a particular estuary rather than at selected point locations. Applications such as these often turn to gridded alternatives to direct observations such as output from directly-observing satellite-based sensors \cite{foreCombinedActivePassive2016} or operational now/fore-casting products \cite{lanerolle2011second}. Unfortunately, these too are of limited use for applications such as validation of next-generation Earth Systems Models (ESMs, \cite{golazDOEE3SMModel2022, xuDevelopmentLandRiver2022a}), where the temporal availability of operational forecasting products is too short ($<$ 1 year) and the spatial resolution of directly-observing satellite-based sensors is too coarse ($>$ 30 $km^2$). \hl{There is a strong need in particular for satellite-based products to validate recently developed biogeochemical modules of the E3SM Earth Systems model which aims to represent thermal, sediment, salinity, nutrient exchanges between rivers and the land on a global basis} \cite{xuDevelopmentLandRiver2022a}.

To address this validation data gap, we developed a data-driven model to produce high-resolution (< 1 $km^2$) estimates of temperature, salinity, and turbidity at daily time-steps (over decadal time scales) as required by next-generation ESMs. The primary input to our model is \hl{remote sensing} reflectance data. In particular, we use Moderate Resolution Imaging Spectroradiometer (MODIS) Aqua multiband data, which has been shown to be predictive of salinity in coastal Louisiana \cite{vogelAssessingSatelliteSea2016, wangDevelopmentMODISData2018}. We supplemented MODIS data with gridded estimates of distance-weighted tributary inflows which we used to encourage physical realism (i.e. geographic “awareness”) especially in areas of the Bay with little to no observations. \hl{Our model was trained against not only buoy observations taken in the mainstem of the Bay but also observations taken in the mid to lower reaches of Bay tributaries.} These more spatially extensive observations likely increase the accuracy and reliability of model predictions in the upper tributaries.

@@

Chesapeake Bay is located in the Mid-Atlantic region of the United States \hl{(at approximately 37.8\degree N 76.1\degree W, see Figure S1)} and serves as the outlet of 9 major tributary systems. Of these, the Susquehanna River (at the Northern extent of the Bay) provides the majority of the freshwater supply to the Bay (\texttildelow$48\%$, \cite{xuClimateForcingSalinity2012}). \hl{The Bay is the largest estuary in the United States and has an area of approximately 11,000 $km^2$. As such, it is provides a wealth of natural economic benefits in the form of fisheries harvests (particularly oysters), tourism, and nature-based recreation} \cite{hoodChesapeakeBayProgram2021}. Despite its large size, the Bay is fairly shallow ($<$ 30m, \cite{murphyComparisonSpatialInterpolation2010}) and the hydrodynamics of the Bay are such that it is considered stratified and partially mixed \cite{xuClimateForcingSalinity2012}. For the present study, this means that surface water properties are likely to differ substantially from water properties at deeper depths. It also means that the Bay receives enough freshwater inputs to produce a distinct upper-estuary to lower-estuary gradient in water properties.

@@

Prior data-driven models (e.g. \cite{urquhartGeospatialInterpolationMODISderived2013, vogelAssessingSatelliteSea2016}) of Bay water properties used Generalized Linear Models (GLM), Neural Networks (NN), Generalized Additive Models (GAMs), and Random Forests (RF). We chose to use the Random Forest algorithm in part because \cite{urquhartGeospatialInterpolationMODISderived2013} found that it produced nearly equivalent results to more complex methods (particularly NN) and also because of its ease of implementation and interpretation. \hl{Random Forests have a particular advantage because they are robust against interactions among variables and allow for nonlinearities} \cite{prasadNewerClassificationRegression2006}. We fit RF models using the scikit-learn Python package \cite{pedregosaScikitlearnMachineLearning2011} where each model was initialized with 250 trees, had a maximum depth of 20 tree “layers”, and was subjected to recursive feature elimination to produce the most parsimonious model (Figure 1). \hl{We chose initial forest size and maximum depth values to optimize model performance and model training time respectively.} We subsequently “tuned” each model using k-fold cross validation to optimize the size, depth, and splitting criteria of each tree. We evaluated the “importance” of particular variables as the average impurity decrease of tree splits based on that variable \cite{pedregosaScikitlearnMachineLearning2011}.

@@

We supplemented observational data with a grid of “freshwater influence” (FWI) estimates where each grid cell represents the average of annual-flow weighted distance of that cell from the 5 major tributaries (Susquehanna, James, Pautexent, Choptank, and Potomac). Annual flow magnitudes were calculated from National Water Information System instantaneous data (sites: 01488110, 01491000, 01576000, 01654000, 01594440, 02037500, \cite{nwis2021usgs}). Rather than straight-line distances we used “in-water” distances calculated using the scikit-image Python package \cite{littleKrigingEstuariesCrow1997,vanderwaltScikitimageImageProcessing2014}. Our in-water distance calculations involved creating a cost–surface whereby land pixels had a very high cost relative to water pixels. Then the distance from any one cell to a tributary source was the accumulated travel “cost” as given by the scikit-image \texttt{route\_through\_array} function. We combined FWI estimates with the pixel-wise daily mean observations and regressed them in RF models against data from the \hl{1000m} resolution MODIS Aqua product, which provides \hl{9} bands of values for each pixel spanning from \hl{the following wavelength ranges (405-420, 438-448, 483-493, 526-536, 546-556, 662-672, 673-683, 743-753, and 862-877nm)} \cite{vermoteericMOD09GAMODISTerra2015}. \hl{MODIS data was pulled from Google Earth Engine (Asset: \texttt{MODIS/006/MYDOCGA}) using the \texttt{earthengine-api} Python package.}

@@

\hl{Across all models, MODIS band 8 was the most important reflectance feature. This result may be partly explained by the fact that Band 8 is the shortest wavelength (405-420nm) included in our study and is likely related to the concentration of Colored Dissolved Organic Matter (CDOM) the absorption of which declines exponentially with increasing wavelength under simulated and laboratory conditions} \cite{loiselleOpticalCharacterizationChromophoric2009, sathyendranath2000remote}). \hl{CDOM itself is strongly related to salinity} \cite{sathyendranath2000remote, vogelAssessingSatelliteSea2016, wangDevelopmentMODISData2018} \hl{and thus may be related to other properties such as temperature and turbidity which also vary spatially with the upstream-downstream environmental gradient of estuaries. In field conditions and in remote sensing applications, the absoption spectra of CDOM and the shape of the CDOM-salinity relationship is highly variable from region-to-region as a function of local factors such as bottom reflectance and phytoplankton community composition} \cite{sathyendranath2000remote}. \hl{Although band 8 was not universally the most important feature in prior remote sensing studies of estuarine water properties, the general importance of short wavelength bands is a result shared by prior studies on estuarine salinity} \cite{wangDevelopmentMODISData2018}.

@@

We have demonstrated an approach to produce realistic estimates of water properties in Chesapeake Bay using \hl{remote sensing} data. Salinity and temperature model performance was as good or better than prior efforts. Salinity model performance was particularly realistic in low salinity tributary areas. Much of this realism comes from our FWI routines which encourage physical consistency in upstream-downstream spatial patterning. We saw additional performance gains owing tothe fact that our training dataset had spatial coverage in the upper tributaries beyond the mainstem of the Bay. Estimates produced by our model have a high potential to inform calibration and evaluation of Earth Systems Models as well as in other applications such as habitat suitability and sea level rise models.

@@

\begin{description}
    \item \textbf{Jemma Stachelek} is a Limnologist and Aquatic Earth Scientist (PhD, Msc, BSc) in the Division of Earth and Environmental Sciences at Los Alamos National Laboratory. Her research focuses on the applications of spatially intensive geocomputation to water and food security problems particularly those related to eutrophication and nutrient pollution.
    \item \textbf{Sofia Avendaño} is awesome.
    \item \textbf{Jon Schwenk} is a River and Data Scientist (PhD) in the Division of Earth and Environmental Sciences at Los Alamos National Laboratory. One of his primary research themes is the use of "big data" as it relates to water to drive models addressing pressing water security issues.    
\end{description}

@@

\begin{description}
    \item Figure 1: Diagram of the data processing, model fitting, tuning, and prediction pipeline.
    \item Figure 2: Out of sample test set performance for salinity, turbidity, and temperature models respectively. Hatching indicates the location of 80\% of the data points.
    \item Figure 3: Random Forest importance plot for salinity, temperature, and turbidity models. \hl{Note that all features may not be present in all models if they were dropped as a result of recursive feature elimination.}
    \item Figure 4: Seasonal Random Forest prediction results for temperature and salinity.
    \item Figure 5: Comparison between Random Forest salinity prediction results and a CBOFS snapshot for Sept, 4, 2022.
\end{description}

>>>>

Unfortunately, direct observations alone are not adequate for many applications that require spatially extensive data such as Sea Level Rise (SLR) forecasting and habitat suitability assessments \cite{hoodChesapeakeBayProgram2021}. For example, habitat suitability assessments are typically done on a gridded basis for the entirety of a particular estuary rather than at selected point locations. Applications such as these often turn to gridded alternatives to direct observations such as output from directly-observing satellite-based sensors \cite{foreCombinedActivePassive2016} or operational now/fore-casting products \cite{lanerolle2011second}. Unfortunately, these too are of limited use for applications such as validation of next-generation Earth Systems Models (ESMs, \cite{golazDOEE3SMModel2022, xuDevelopmentLandRiver2022a}), where the temporal availability of operational forecasting products is too short ($<$ 1 year) and the spatial resolution of directly-observing satellite-based sensors is too coarse ($>$ 30 $km^2$). There is a strong need in particular for satellite-based products to validate recently developed biogeochemical modules of the E3SM Earth Systems model which aims to represent thermal, sediment, salinity, nutrient exchanges between rivers and the land on a global basis \cite{xuDevelopmentLandRiver2022a}.

To address this validation data gap, we developed a data-driven model to produce high-resolution (< 1 $km^2$) estimates of temperature, salinity, and turbidity at daily time-steps (over decadal time scales) as required by next-generation ESMs. The primary input to our model is remote sensing reflectance data. In particular, we use Moderate Resolution Imaging Spectroradiometer (MODIS) Aqua multiband data, which has been shown to be predictive of salinity in coastal Louisiana \cite{vogelAssessingSatelliteSea2016, wangDevelopmentMODISData2018}. We supplemented MODIS data with gridded estimates of distance-weighted tributary inflows which we used to encourage physical realism (i.e. geographic “awareness”) especially in areas of the Bay with little to no observations. Our model was trained against not only buoy observations taken in the mainstem of the Bay but also observations taken in the mid to lower reaches of Bay tributaries. These more spatially extensive observations likely increase the accuracy and reliability of model predictions in the upper tributaries.

@@

Chesapeake Bay is located in the Mid-Atlantic region of the United States (at approximately 37.8\degree N 76.1\degree W, see Figure S1) and serves as the outlet of 9 major tributary systems. Of these, the Susquehanna River (at the Northern extent of the Bay) provides the majority of the freshwater supply to the Bay (\texttildelow$48\%$, \cite{xuClimateForcingSalinity2012}). The Bay is the largest estuary in the United States and has an area of approximately 11,000 $km^2$. As such, it is provides a wealth of natural economic benefits in the form of fisheries harvests (particularly oysters), tourism, and nature-based recreation \cite{hoodChesapeakeBayProgram2021}. Despite its large size, the Bay is fairly shallow ($<$ 30m, \cite{murphyComparisonSpatialInterpolation2010}) and the hydrodynamics of the Bay are such that it is considered stratified and partially mixed \cite{xuClimateForcingSalinity2012}. For the present study, this means that surface water properties are likely to differ substantially from water properties at deeper depths. It also means that the Bay receives enough freshwater inputs to produce a distinct upper-estuary to lower-estuary gradient in water properties.

@@

Prior data-driven models (e.g. \cite{urquhartGeospatialInterpolationMODISderived2013, vogelAssessingSatelliteSea2016}) of Bay water properties used Generalized Linear Models (GLM), Neural Networks (NN), Generalized Additive Models (GAMs), and Random Forests (RF). We chose to use the Random Forest algorithm in part because \cite{urquhartGeospatialInterpolationMODISderived2013} found that it produced nearly equivalent results to more complex methods (particularly NN) and also because of its ease of implementation and interpretation. Random Forests have a particular advantage because they are robust against interactions among variables and allow for nonlinearities \cite{prasadNewerClassificationRegression2006}. We fit RF models using the scikit-learn Python package \cite{pedregosaScikitlearnMachineLearning2011} where each model was initialized with 250 trees, had a maximum depth of 20 tree “layers”, and was subjected to recursive feature elimination to produce the most parsimonious model (Figure 1). We chose initial forest size and maximum depth values to optimize model performance and model training time respectively. We subsequently “tuned” each model using k-fold cross validation to optimize the size, depth, and splitting criteria of each tree. We evaluated the “importance” of particular variables as the average impurity decrease of tree splits based on that variable \cite{pedregosaScikitlearnMachineLearning2011}.

@@

We supplemented observational data with a grid of “freshwater influence” (FWI) estimates where each grid cell represents the average of annual-flow weighted distance of that cell from the 5 major tributaries (Susquehanna, James, Pautexent, Choptank, and Potomac). Annual flow magnitudes were calculated from National Water Information System instantaneous data (sites: 01488110, 01491000, 01576000, 01654000, 01594440, 02037500, \cite{nwis2021usgs}). Rather than straight-line distances we used “in-water” distances calculated using the scikit-image Python package \cite{littleKrigingEstuariesCrow1997,vanderwaltScikitimageImageProcessing2014}. Our in-water distance calculations involved creating a cost–surface whereby land pixels had a very high cost relative to water pixels. Then the distance from any one cell to a tributary source was the accumulated travel “cost” as given by the scikit-image \texttt{route\_through\_array} function. We combined FWI estimates with the pixel-wise daily mean observations and regressed them in RF models against data from the 1000m resolution MODIS Aqua product, which provides 9 bands of values for each pixel spanning from the following wavelength ranges (405-420, 438-448, 483-493, 526-536, 546-556, 662-672, 673-683, 743-753, and 862-877nm) \cite{vermoteericMOD09GAMODISTerra2015}. MODIS data was pulled from Google Earth Engine (Asset: \texttt{MODIS/006/MYDOCGA}) using the \texttt{earthengine-api} Python package.

@@

Across all models, MODIS band 8 was the most important reflectance feature. This result may be partly explained by the fact that Band 8 is the shortest wavelength (405-420nm) included in our study and is likely related to the concentration of Colored Dissolved Organic Matter (CDOM) the absorption of which declines exponentially with increasing wavelength under simulated and laboratory conditions \cite{loiselleOpticalCharacterizationChromophoric2009, sathyendranath2000remote}. CDOM itself is strongly related to salinity \cite{sathyendranath2000remote, vogelAssessingSatelliteSea2016, wangDevelopmentMODISData2018} and thus may be related to other properties such as temperature and turbidity which also vary spatially with the upstream-downstream environmental gradient of estuaries. In field conditions and in remote sensing applications, the absoption spectra of CDOM and the shape of the CDOM-salinity relationship is highly variable from region-to-region as a function of local factors such as bottom reflectance and phytoplankton community composition \cite{sathyendranath2000remote}. Although band 8 was not universally the most important feature in prior remote sensing studies of estuarine water properties, the general importance of short wavelength bands is a result shared by prior studies on estuarine salinity \cite{wangDevelopmentMODISData2018}.

@@

We have demonstrated an approach to produce realistic estimates of water properties in Chesapeake Bay using remote sensing data. Salinity and temperature model performance was as good or better than prior efforts. Salinity model performance was particularly realistic in low salinity tributary areas. Much of this realism comes from our FWI routines which encourage physical consistency in upstream-downstream spatial patterning. We saw additional performance gains owing tothe fact that our training dataset had spatial coverage in the upper tributaries beyond the mainstem of the Bay. Estimates produced by our model have a high potential to inform calibration and evaluation of Earth Systems Models as well as in other applications such as habitat suitability and sea level rise models.

@@

\begin{description}
    \item \textbf{Jemma Stachelek} is a Limnologist and Aquatic Earth Scientist (PhD, Msc, BSc) in the Division of Earth and Environmental Sciences at Los Alamos National Laboratory. Her research focuses on the applications of spatially intensive geocomputation to water and food security problems particularly those related to eutrophication and nutrient pollution.
    \item \textbf{Sofia Avendaño} is a hydrogeologist with a Master's from New Mexico Institute of Mining and Technology. She has worked on a wide variety of projects including fracture flow, surface flow and hydrodynamics. Her scientific passion is modeling fluid flow and sediment transport.
    \item \textbf{Jon Schwenk} is a River and Data Scientist (PhD) in the Division of Earth and Environmental Sciences at Los Alamos National Laboratory. One of his primary research themes is the use of "big data" as it relates to water to drive models addressing pressing water security issues.    
\end{description}

@@

\begin{description}
    \item Figure 1: Diagram of the data processing, model fitting, tuning, and prediction pipeline.
    \item Figure 2: Out of sample test set performance for salinity, turbidity, and temperature models respectively. Hatching indicates the location of 80\% of the data points.
    \item Figure 3: Random Forest importance plot for salinity, temperature, and turbidity models. Note that all features may not be present in all models if they were dropped as a result of recursive feature elimination.
    \item Figure 4: Seasonal Random Forest prediction results for temperature and salinity.
    \item Figure 5: Comparison between Random Forest salinity prediction results and a CBOFS snapshot for Sept, 4, 2022.
\end{description}
